# Land-Cover classification {#sec-land-cover}

In previous sections I've explained how seasonality in vegetation canopy density (LAI) or phenology can be detected, and how it varies depending on the geography of the region, and its ties to the exchange of carbon (C) and water between the biosphere and the atmosphere (@sec-phenology-trends). A small, first principles, example was provided on how to write your own phenology detection algorithm (@sec-algorithms).

However, many satellite platforms provide information in various spectral bands (and not only one product in the temporal domain). One can say the data are multi-dimensional, having both a temporal, spatial and spectral component. These various bands, or locations in the spectral domain, provide key insights into the state of the land surface (throughout the year).

For example the combination of information in the red and near-infrared bands (spectral domains) provides key information to calculate the Normalized Difference Vegetation Index [@Huete2002]. Other band combinations lead to other indices with varying properties, tailored to specific ecological, geomorphological or other purposes [@zeng_optical_2022]. 

When we plot a time serie of a deciduous forest you note the steady seasonal changes when switching between winter, with low NDVI values, and summer with high NDVI values. However, different vegetation and land cover types have different temporal NDVI signals. For example, a glacier will have permanent snow and no seasonal NDVI signal. We can therefore discriminate between non-vegetative locations and vegetation based upon the temporal profile of NDVI.

```{r}
#| label: fig-ndvi-dbf
#| fig-cap: "The mean and standard deviation of the normalized difference vegetation index (NDVI) over a broadleaf forest"
#| fig-align: "center"
#| out-width: "100%"
#| echo: FALSE

```


```{r}
#| label: fig-ndvi-ice
#| fig-cap: "The mean and standard deviation of the normalized difference vegetation index (NDVI) over a glacier"
#| fig-align: "center"
#| out-width: "100%"
#| echo: FALSE


```

As such we can use this information to classify the Swiss alpine scene into locations which have little seasonality and those which have some (e.g. @fig-ndvi-dbf). For example you can calculate the mean and standard deviation of a full year and see how much variability you see across a year. Regions with a low NDVI signal with little variability are most likely not associated with vegetation (e.g. lakes, glaciers, rocky surfaces, urban areas, see @fig-ndvi-ice).

However, a simple index as an NDVI alone does not provide sufficient information to distinguish between more subtle vegetation or land-cover classes (e.g. evergreen forests and or mixed forest types). Instead we can use individual spectral bands (some of which make up the NDVI index) to create a more detailed classification of land cover in different (vegetation) types.

## (supervised) Machine learning

Classification of data in different classes (or clustering) can be accomplished using various methods. Clustering can either be unsupervised, where clustering is only defined by the number of classes one wants to divide the (multi-dimensional) dataset into. In contrast, one can use supervised classification of data where reference data is used as a benchmark on which to train a machine learning algorithm. In this supervised machine learning an algorithm will try to minimize the classification error on a known training or validation dataset. 

In our case the temporal (seasonal) shape of the spectral data would be used to characterize different land-cover types.

```{r}
#| label: fig-spectral-bands
#| fig-cap: "The shape of a single deciduous broadleaf pixel across various MODIS bands"
#| fig-align: "center"
#| out-width: "100%"
#| echo: FALSE


```

A full introduction to machine learning beyond the scope of this course and I refer to XYZ and XYZ for a (free) introduction in machine learning using R.

### Land-cover validation data

Critical in our exercise is the reference data we use to classify different land-cover types. Generally ground truth data is used in order to create a training dataset. This ground truth data are locations which are visually validated to belong to a paurticular land-cover class.

These data are gathered by *in-situ* surveys, or leverage high(er) resolution data to confirm the the land-cover type. For example, the LUCAS database of the European Commission provides 3 yearly survey data on the land-cover state (including detailed photographs of the surrounding of a validation point).

WHAT DOES MODIS MCD12Q1 USE?

Such up to date training and validation data is crucial in the development of supervised land-cover maps. In the below worked example we will use the

## Multi-spectral data to characterize land-cover

https://rspatial.org/terra/rs/5-supclassification.html
